---
title: "Introduction to R for Data Exploration and Plotting"
subtitle: "Answer key"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
---

## Setup

```{r}
data(iris)
library(tidyverse)
iris_df <- iris
```

## Exercise 1

Create a vector called `ages` containing five ages. Then calculate the mean age.

```{r}
ages <- c(18, 21, 25, 30, 34)
mean(ages)
```

## Exercise 2

Using base R:

1. Show only rows where `Species` is `versicolor`.
2. Show only the columns `Sepal.Length`, `Petal.Length`, and `Species`.

```{r}
iris[iris$Species == "versicolor", ]

iris[, c("Sepal.Length", "Petal.Length", "Species")]
```

## Exercise 3

Using tidyverse, calculate the mean `Sepal.Length` and mean `Petal.Length` for each species.

```{r}
iris_df %>%
  dplyr::group_by(Species) %>%
  dplyr::summarise(
    mean_sepal_length = mean(Sepal.Length),
    mean_petal_length = mean(Petal.Length)
  )
```

## Exercise 4

Make a scatter plot with:

1. `Sepal.Length` on the x-axis.
2. `Sepal.Width` on the y-axis.
3. Points colored by `Species`.
4. A clean theme using `theme_bw()`.
5. Clear axis labels.

```{r}
ggplot(data = iris_df, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 3) +
  labs(
    x = "Sepal length",
    y = "Sepal width",
    color = "Species"
  ) +
  theme_bw()
```

## Final challenge

Complete the following workflow:

1. Remove `setosa`.
2. Create a new column called `Petal.Ratio`, calculated as `Petal.Length / Petal.Width`.
3. Calculate the mean `Petal.Ratio` for each species.
4. Make a boxplot of `Petal.Ratio` by species.

```{r}
iris_df2 <- iris_df %>%
  dplyr::filter(Species != "setosa") %>%
  dplyr::mutate(
    Petal.Ratio = Petal.Length / Petal.Width
  )

iris_df2 %>%
  dplyr::group_by(Species) %>%
  dplyr::summarise(
    mean_petal_ratio = mean(Petal.Ratio)
  )

ggplot(iris_df2, aes(x = Species, y = Petal.Ratio, fill = Species)) +
  geom_boxplot() +
  labs(
    x = "Species",
    y = "Petal length / petal width"
  ) +
  theme_bw()
```
