---
title: "HAAM Summer School – Introduction to R"
author: "Youssef Tawfik"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: true
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

## Workshop goals

By the end of this session, you should be able to:

1.  Run simple R commands in RStudio.
2.  Understand objects, functions, vectors, and data frames.
3.  Inspect and subset a dataset using base R.
4.  Use basic `dplyr` commands to filter, select, mutate, group, and summarise data.
5.  Create simple plots using `ggplot2`.
6.  Save a plot as an image file.

We will use the built-in `iris` dataset, so no external data files are needed.

## 1. Getting started with R

R can be used like a calculator.

```{r}
2 + 2
10 / 2
sqrt(25)
```

We can save values as objects using `<-`.

```{r}
x <- 10
x

x + 5
```

A vector is a collection of values.

```{r}
numbers <- c(1, 2, 3, 4, 5)
numbers

mean(numbers)
sd(numbers)
length(numbers)
```

### Exercise 1

Create a vector called `ages` containing five ages. Then calculate the mean age.

```{r}
# Write your code here

```

## 2. Working with a data frame

A data frame is a table. Rows are observations and columns are variables.

We will use the built-in `iris` dataset.

```{r}
data(iris)
```

Inspect the dataset.

```{r}
head(iris)
str(iris)
summary(iris)
dim(iris)
colnames(iris)
```

The dataset has measurements from three iris species.

```{r}
unique(iris$Species)
```

We can access one column using `$`.

```{r}
iris$Petal.Length
```

We can subset rows and columns using square brackets.

```{r}
# First five rows
iris[1:5, ]

# Rows where Species is setosa
iris[iris$Species == "setosa", ]

# Rows where Petal.Length is greater than 5
iris[iris$Petal.Length > 5, ]

# Select specific columns
iris[, c("Species", "Petal.Length", "Petal.Width")]
```

### Exercise 2

Using base R:

1.  Show only rows where `Species` is `versicolor`.
2.  Show only the columns `Sepal.Length`, `Petal.Length`, and `Species`.

```{r}
# Write your code here

```

## 3. Data manipulation with tidyverse

The tidyverse is a collection of R packages for data manipulation and plotting.

```{r}
library(tidyverse)
```

For this workshop, we will copy `iris` into an object called `iris_df`.

```{r}
iris_df <- iris
```

The pipe `%>%` can be read as "then".

```{r}
iris_df %>%
  head()
```

### Filtering rows

```{r}
iris_df %>%
  dplyr::filter(Species == "setosa")
```

```{r}
iris_df %>%
  dplyr::filter(Petal.Length > 5)
```

### Selecting columns

```{r}
iris_df %>%
  dplyr::select(Species, Petal.Length, Petal.Width)
```

### Creating new columns

```{r}
iris_df %>%
  dplyr::mutate(
    Petal.Ratio = Petal.Length / Petal.Width
  )
```

### Grouping and summarising

```{r}
iris_df %>%
  dplyr::group_by(Species) %>%
  dplyr::summarise(
    mean_petal_length = mean(Petal.Length),
    mean_petal_width = mean(Petal.Width),
    n = dplyr::n()
  )
```

### Exercise 3

Using tidyverse, calculate the mean `Sepal.Length` and mean `Petal.Length` for each species.

```{r}
# Write your code here

```

## 4. Plotting with ggplot2

`ggplot2` builds plots in layers.

The basic structure is:

``` r
/* This block is shown as text only. Do not run it. */
ggplot(data, aes(x = variable1, y = variable2)) +
  geom_something()
```

### A basic scatter plot

```{r}
ggplot(data = iris_df, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point()
```

### Color points by species

```{r}
ggplot(data = iris_df, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point()
```

### Improve the plot appearance

```{r}
ggplot(data = iris_df, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "Petal measurements in iris species",
    x = "Petal length",
    y = "Petal width",
    color = "Species"
  ) +
  theme_bw()
```

### A boxplot

```{r}
ggplot(data = iris_df, aes(x = Species, y = Petal.Length, fill = Species)) +
  geom_boxplot() +
  labs(
    title = "Petal length differs between iris species",
    x = "Species",
    y = "Petal length"
  ) +
  theme_bw()
```

### Combining tidyverse and ggplot2

```{r}
iris_df %>%
  dplyr::filter(Species != "setosa") %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point(size = 3) +
  theme_bw()
```

### Exercise 4

Make a scatter plot with:

1.  `Sepal.Length` on the x-axis.
2.  `Sepal.Width` on the y-axis.
3.  Points colored by `Species`.
4.  A clean theme using `theme_bw()`.
5.  Clear axis labels.

```{r}
# Write your code here

```

## 5. Saving a plot

First, save the plot as an object.

```{r}
p <- ggplot(data = iris_df, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "Petal measurements in iris species",
    x = "Petal length",
    y = "Petal width",
    color = "Species"
  ) +
  theme_bw()

p
```

Then save it with `ggsave()`.

```{r}
ggsave("iris_petal_plot.png", p, width = 6, height = 4, dpi = 300)
```

## 6. Final challenge

Complete the following workflow:

1.  Remove `setosa`.
2.  Create a new column called `Petal.Ratio`, calculated as `Petal.Length / Petal.Width`.
3.  Calculate the mean `Petal.Ratio` for each species.
4.  Make a boxplot of `Petal.Ratio` by species.

```{r}
# Write your code here

```

## 7. Key take-home messages

-   R stores information in objects.
-   Most R work involves creating, modifying, and plotting data frames.
-   Base R is useful for understanding the structure of R.
-   Tidyverse makes many data manipulation tasks easier to read.
-   `ggplot2` builds figures by adding layers.
-   You do not need to memorize everything. You need to understand the structure and know how to adapt examples.
