---
title: "HAAM Summer School – Introduction to R"
author: "Youssef Tawfik"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: false
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

## 1. Objectives

By the end of this session, you should be able to:

1.  Run simple R commands in RStudio.
2.  Understand objects, functions, vectors, and data frames.
3.  Inspect and subset a dataset using base R.
4.  Read (load) and write (save) files/datasets.
5.  Use basic `dplyr` commands to filter and subset data.
6.  Create plots using `ggplot2`.
7.  Save a plot as an image file.

## 2. Getting started with R & RStudio

### What is R?

R is a language and environment for statistical programming which includes statistical computing and graphics.

### RStudio

It is an integrated development environment (IDE) for the R programming language.

*We will now open RStudio and learn how to navigate it and use R in it.*

### Working in RStudio

R can be used like a calculator.

```{r}
#| output: false
2 + 2
10 / 2
sqrt(25)
```

We can save values as objects using `<-`.

```{r}
#| output: false
x <- 10
x

x + 5
```

A vector is a collection of values.

```{r}
#| output: false
numbers <- c(1, 2, 3, 4, 5)
numbers
```

A function is a command with input(s).

```{r}
#| output: false
mean(numbers)
sd(numbers)
length(numbers)
```

You can check what a function does, what kind of input it requires, and what kind of output it produces using `?`

```{r}
#| output: false
?mean
```

### Exercise 1

Now, create a vector called `ages` containing the values: 18, 22, 26, 28, 32. Calculate the mean age.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| output: false

ages <- c(18, 22, 26, 28, 32)
mean(ages) # = 25.2

```

## 3. Working with a data frame

You can think of a data frame as a table with rows (observations) and columns (variables).

We will use the built-in `iris` dataset. We will first load it:

```{r}
#| output: false

data(iris)
```

Now, let's inspect the dataset.

```{r}
#| output: false

# show the first few lines
head(iris)
# show structure of the data frame
str(iris)
# show summary of the data frame
summary(iris)
# show dimmensions of the data frame
dim(iris)
# show the column names of the data frame
colnames(iris)
```

The dataset has measurements from three iris species.

```{r}
#| output: false

unique(iris$Species)
```

We can access one column using `$`.

```{r}
#| output: false

iris$Petal.Length
```

We can subset rows and columns using square brackets.

```{r}
#| output: false

# First five rows
iris[1:5, ]

# Rows where Species is setosa
iris[iris$Species == "setosa", ]

# Rows where Petal.Length is greater than 5
iris[iris$Petal.Length > 5, ]

# Select specific columns
iris[, c("Species", "Petal.Length", "Petal.Width")]
```

We can also add new columns to our data frame.

```{r}
#| output: false

iris[,6] <- iris$Petal.Length / iris$Petal.Width
```

We can check and/or change column names of a data frame using `colnames()`.

```{r}
#| output: false

colnames(iris)

colnames(iris)[6] <- "Petal.Ratio"
```

### Exercise 2

1.  Show only rows where `Species` is `versicolor`.
```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| output: false

iris[iris$Species == "versicolor", ]

```

2.  Show only the columns `Sepal.Length`, `Petal.Length`, and `Species`.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| output: false

iris[, c("Sepal.Length", "Petal.Length", "Species")]

```

## 4. Reading (loading) and writing (saving) files

When using R, we will always have input files that need to be loaded and/or output files that need to be saved.

We will copy `iris` into an object called `iris_df`.

```{r}
#| output: false

iris_df <- iris
```

You should always know where you are. We use `getwd()` to know current directory (folder) we are in.

We use `setwd()` to set a new working directory (folder) to work in.

```{r}
#| eval: false
setwd("/vol/volume/intro-to-r/")
```

We can check files inside current working directory using `list.files()`

Let's now save the `iris` data frame into a text file.

```{r}
#| eval: false

write_tsv(x = iris_df, file = "iris_df.tsv")
```

Now, let's remove the current *iris_df* and try to read (load) it from our working directory.

```{r}
#| eval: false

rm(iris_df)
iris_df <- read_tsv("iris_df.tsv")
```


## 5. Data manipulation with tidyverse

The tidyverse is a collection of R packages for data manipulation and plotting.

```{r}
library(tidyverse)
```

Let's again print the first lines of the iris data frame.

```{r}
#| output: false

head(iris_df)
```

We can also use the pipe `%>%` (which can be read as "then"). The syntax is simple: *object* `%>%` *function*

```{r}
#| output: false

iris_df %>%
  head()
```

### Filtering rows

We use `dplyr::filter()` and indicate what we want to subset our data frame with. 

```{r}
#| output: false

# Filter by specific value
iris_df %>%
  dplyr::filter(Species == "setosa")
# Filter by range
iris_df %>%
  dplyr::filter(Petal.Length > 5)
```

### Selecting specific columns

We use `dplyr::select()` and indicate which columns we want to subset from our data frame.

```{r}
#| output: false

iris_df %>%
  dplyr::select(Species, Petal.Length, Petal.Width)
```

### Creating new columns

We can also use `dplyr::mutate()` to create new columns.

```{r}
#| output: false

iris_df %>%
  dplyr::mutate(
    Petal.Ratio = Petal.Length / Petal.Width
  )
```

### Exercise 4

What is the difference between these two codes?

```{r}
#| output: false

# code 1:
iris[,6] <- iris$Petal.Length / iris$Petal.Width
colnames(iris)

colnames(iris)[6] <- "Petal.Ratio"

# code 2:
iris_df %>%
  dplyr::mutate(
    Petal.Ratio = Petal.Length / Petal.Width
    )

```

## 6. Plotting with ggplot2

`ggplot2` builds plots in layers.

The basic structure is:


*ggplot(data, aes(x = variable1, y = variable2)) +*
  *geom_something()*

### A basic scatter plot

```{r}
ggplot(data = iris_df, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point()
```

### Color points by species

```{r}
ggplot(data = iris_df, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point()
```

### Improve the plot appearance

```{r}
ggplot(data = iris_df, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "Petal measurements in iris species",
    x = "Petal length",
    y = "Petal width",
    color = "Species"
  ) +
  theme_bw()
```

### A boxplot

```{r}
ggplot(data = iris_df, aes(x = Species, y = Petal.Length, fill = Species)) +
  geom_boxplot() +
  labs(
    title = "Petal length differs between iris species",
    x = "Species",
    y = "Petal length"
  ) +
  theme_bw()
```

### Combining dplyr and ggplot2

We can plot a subset of a data frame by using `dplyr` and `ggplot2` together (without creating a new data frame).

```{r}
iris_df %>%
  dplyr::filter(Species != "setosa") %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point(size = 3) +
  theme_bw()
```

### Exercise 5

Make a scatter plot with:

1.  `Sepal.Length` on the x-axis.
2.  `Sepal.Width` on the y-axis.
3.  Points colored by `Species`.
4.  A clean theme using `theme_bw()`.
5.  Clear axis labels.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

ggplot(data = iris_df, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 3) +
  labs(
    x = "Sepal length",
    y = "Sepal width",
    color = "Species"
  ) +
  theme_bw()

```

## 7. Saving a plot

Finally, we want to save our plot.

To do that, we can first save the plot as an object.

```{r}
p <- ggplot(data = iris_df, aes(x = Petal.Length, y = Petal.Width, color = Species)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "Petal measurements in iris species",
    x = "Petal length",
    y = "Petal width",
    color = "Species"
  ) +
  theme_bw()

p
```

Then, we can save it using `ggsave()`.

```{r}
#| eval: false
ggsave(filename = "iris_petal_plot.png", plot = p)
```

## 8. Resources

-   [Posit Cheat Sheets](https://rstudio.github.io/cheatsheets/)  
-   [RStudio / Posit Cheat Sheets collection](https://opensource.posit.co/resources/cheatsheets/) 
-   [R Graph Gallery cheat sheets](https://r-graph-gallery.com/cheatsheets)  
-   [Stack Overflow](https://stackoverflow.com)  
-   [Babraham Bioinformatics courses](https://www.bioinformatics.babraham.ac.uk/training.html)  
-   [Statistical Tools for High-throughput Data Analysis](sthda.com/english/)
-   [Base-R cheat sheet](https://iqss.github.io/dss-workshops/R/Rintro/base-r-cheat-sheet.pdf)

